Journal and article coverage

### 2025-03-18 analysis/manuscript.tex
Some minor language fixes

This study investigates the suitability of open scholarly data sources
for assessing the impact of transformative agreements on hybrid open
access. As shown in Figure \ref{fig:workflow}, the methodology involved
comparing hoaddata, an openly available collection of open research
information on hybrid open access, with the bibliometric databases Web
of Science and Scopus. This section introduces the initial data sources,
followed by a presentation of the necessary data processing steps to
obtain eligible articles from transformative agreements by using author
roles (first and corresponding) and harmonised affiliation data.

@@

hoaddata, developed and maintained by the author to support open access
monitoring and research (Jahn, 2025), is an R data package collecting
information on hybrid open access uptake from multiple openly available
data sources. It combines article-level metadata from Crossref and
OpenAlex with transformative agreement information from the cOAlition S
Journal Checker Tool (JCT), which links journal and institutional data
to agreements in the ESAC registry.

As discussed elsewhere (Jahn, 2025), hoaddata uses a unified dataset of
historic JCT snapshots, which were safeguarded since July 2021. Journal
data were enriched with ISSN variants linked to an ISSN-L and insitution
data with associated ROR-IDs representing university hospitals or
institutes of large research organisations, according to OpenAlex'
institution entity. A comprehensive exclusion of fully open access
journals using multiple journal lists (DOAJ, OpenAlex, Bielefeld) was
performed to identify hybrid journals. hoaddata relies on Crossref for
obtaining journal publication volume and open access status through
Creative Commons licence information relative to the published version
(``version of record''). Because of limited affiliation metadata in
Crossref (Eck \& Waltman, 2022), hoaddata sources affiliation metadata
from OpenAlex.

hoaddata follows good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow handles
data retrieval from the SUB Göttingen's open scholarly data warehouse
based on Google BigQuery, which provides high-performant programmatic
access to monthly snapshots of Crossref and OpenAlex. The workflow has
run regularly to fetch updates from these data sources since 2022. The
package version used in this study is 0.3, containing data from the
Crossref 2024-08 dump provided to Crossref Metadata Plus subscribers and
the OpenAlex 2024-08-29 monthly dump. This version including the
computation log is available on GitHub
(https://github.com/subugoe/hoaddata/releases/tag/v.0.3).

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). The collections considered in this study were the
Science Citation Index Expanded (SCIE), the Social Sciences Citation
Index (SSCI) and the Arts \& Humanities Citation Index (AHCI),
collectively referred to as Web of Science Primary (WoS Primary).

The WoS Primary provides important data points for analysing open
access: author affiliations and roles, differentiation of journal
articles into document types representing different types of journal
contributions, such as original articles or reviews, and open access
status information derived from OurResearch's Unpaywall, the same
provider as OpenAlex. However, it lacks information about journals and
articles under transformative agreements.

For programmatic access to article-level data, this study used the
database of the Kompetenznetzwerk Bibliometrie (KB) in Germany. The KB
processes raw XML data provided by Clarivate Analytics, which is
provided as an in-house PostgreSQL database under a uniform schema. To
support reproducibility, KB maintains annual snapshots of the database.
Accordingly, this study used the annual snapshot from April 2024
(wos\_b\_202404), which is considered to cover almost the entire
previous publication year (Schmidt et al., 2024).

@@

The hoaddata package (v.0.3) provided an enriched and unified list of
hybrid journals in transformative agreements according to the JCT,
covering agreements active from July 2021 to July 2024 according to the
ESAC registry.

Article metadata were retrieved from each database using all ISSN
variants linked to an ISSN-L. The metadata included DOIs, publication
dates, article types, open access information as well as author roles
and institutional affiliations. Publication years were determined using
the earliest known date of publication in a journal. In hoaddata, this
corresponded to Crossref's issued date. For Web of Science and Scopus,
the earliest publication date was used where available, with Scopus
dates specifically determined by the KB through version tracking of the
raw data.

Many transformative agreements typically cover only certain types of
journal articles, in particular original research articles including
reviews. Because of limited information on these document types in open
scholarly data (Haupka et al., 2024), hoaddata used an extended version
of Unpaywall's paratext recognition approach to exclude non-scholarly
content (Piwowar et al., 2018). To exclude conference supplements, which
are also often not covered by transformative agreements, only articles
published in regular issues, indicated by numerical pagination, were
considered. For Web of Science and Scopus, their established document
type classifications were used to identify original research articles
and reviews, referred to as core articles throughout this study.

@@

Published articles were considered hybrid open access when they were
made freely available under a Creative Commons license on publishers'
platforms. While hoaddata sourced this information from Crossref license
metadata, Web of Science and Scopus relied on Unpaywall. Unpaywall
supplements Crossref metadata by parsing publisher websites directly,
addressing cases where publishers do not provide machine-readable
Creative Commons license information (Piwowar et al., 2018). This
additional parsing remains necessary despite transformative agreement
workflows recommend the depositing license information during DOI
registration (Geschuhn \& Stone, 2017). Both Web of Science and Scopus
defined hybrid open access consistently as content available under
Creative Commons licenses on publisher platforms, distinguishing it from
bronze open access that lack such explicit license information.

@@

Author affiliations were retrieved for both first and, if available,
corresponding authors across all databases to prepare the linking
between articles and institutions covered by transformative agreements.
To handle different address variants, database-specific affiliation
identifiers were used: ROR-IDs from OpenAlex for hoaddata, affiliation
enhanced names for Web of Science, and Scopus Affiliation Identifiers
for Scopus. Additionally, ISO country codes were retrieved for each
author's address to compile country-level statistics. These country
codes for Web of Science and Scopus were provided by the KB.

Because neither Web of Science nor Scopus support ROR IDs---the
institution identifier used by the JCT---a two-step matching process was
implemented to harmonise affiliation data. First, 2,782,540 articles
from 6,457 institutions with ROR-IDs in the JCT data since 2017
(according to hoaddata) were processed to map first authors' ROR-IDs to
corresponding proprietary affiliation identifier in Web of Science and
Scopus using DOI matching. Then, an algorithm selected the most frequent
ROR ID and proprietary identifier pairs to handle multiple affiliations
and organisational hierarchy differences.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were aggregated for each database at country and journal
level by year. Table~\ref{tbl-methods_overview_tab} provides a general
overview of the coverage between 2019 and 2023 per database. It shows
that the majority of hybrid journals published at least one original
research article or review marked as core during the five-year period.
These journals formed the basis for the subsequent calculation of
article-level indicators.

@@

\centering
\begin{tabular}[t]{lrrr}
\toprule
\textbf{} & \textbf{HOAD} & \textbf{Web of Science} & \textbf{Scopus}\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Hybrid journal coverage}}\\
\hspace{1em}Active journals & 12,890 & 8,655 & 11,888\\
\hspace{1em}Active journals (core) & 12,888 & 8,655 & 11,878\\
\hspace{1em}Active journals (core) with OA & 11,348 & 8,392 & 11,313\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Publication volume}}\\
\hspace{1em}Total published articles & 9,740,015 & 8,616,053 & 8,117,644\\
\hspace{1em}Core articles & 8,158,425 & 6,708,083 & 7,317,703\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Digital Object Identifier (DOI) coverage}}\\
\hspace{1em}Articles with DOI & 9,740,015 & 7,713,796 & 8,105,112\\
\hspace{1em}Core articles with DOI & 8,158,425 & 6,695,661 & 7,314,327\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Open Access (OA) metrics}}\\
\hspace{1em}OA articles & 998,699 & 1,112,758 & 974,099\\
\hspace{1em}Core OA articles & 969,817 & 1,019,784 & 922,578\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Core articles with affiliation data}}\\
\hspace{1em}First author articles & 7,242,542 & 6,294,855 & 7,232,017\\
\hspace{1em}Corresponding author articles & 5,534,207 & 6,291,441 & 6,898,487\\
\bottomrule
\end{tabular}

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that non-core
articles in Web of Science often lacked a DOI. This was particularly the
case for meeting abstracts, which are notably prevalent in Health
Sciences journals (Melero-Fuentes et al., 2025) and are not indexed by
Scopus (Donner, 2017). Open access indicators were aggregated by DOI, as
Unpaywall only collects information on open access status for articles
with a DOI. A closer look at the core articles with affiliation data
reveals a lack of data on corresponding authors in the case of OpenAlex,
the affiliation data source used by hoaddata, compared to Web of Science
and Scopus. Only about two-thirds of the articles examined provided
corresponding author affiliations. For first authors, the proportion was
89\%. Therefore, only first author data for hoaddata were considered in
the following analysis.

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second-largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaadata, while 6\% (n = 748)
were only found in the proprietory database Scopus. Of these, 354 were
also available in the Web of Science. Upon inspection, this group of
hybrid journals included only in the proprietary data sources mainly
represents hybrid journals for which no open access evidence could be
retrieved from Crossref, which served as the open access evidence source
for hoaddata.

In terms of article coverage, Figure \ref{fig-upset_coverage_results}B
shows the total publication volume per combination in terms of DOI
availability. The largest set of hybrid journals, which includes all
three data sources, also contains the largest number of articles. In
total, these journals recorded 6,289,687 overlapping articles,
represented by the blue bar. They represent 94\% of core articles with
DOI indexed in the Web of Science sample, and 86\% in Scopus. Another
657,697 articles were exclusive to both Scopus and hoaddata. Exclusively
in hoaddata were 177,110 articles, and exclusively in the proprietary
databases were 325,194 articles.

Figure \ref{fig-upset_coverage_results}B also shows DOIs that were only
available via hoaddata (grey area). In case of hybrid journals covered
by all three data sources, 1,023,882 DOIs were only present in hoaddata.
After validation at the DOI level using the KB databases and manual
inspection, the main reasons for missing DOI coverage in the proprietary
database were insufficient classification of journal content as original
research articles and reviews during the compilation of hoaddata.
Particulalry, letters and editorials could not be fully detected.
Paratext recognition failed in 37\% of DOIs to identify non-scientific
content such as front matters or reviewer lists, which are generally not
indexed by Scopus and Web of Science. To a lesser extent, differences in
publication and indexing dates were a reason for non-overlapping DOIs.

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows that there is a large spread across the journals
covered by all three data sources. Furthermore, journals in this set
published more on average than journals covered by only one or two data
sources. In particular, the journals covered exclusively by hoaddata are
substantially smaller than those covered by all three sources. Upon
inspection, these are often newly launched journals, which explains the
relatively low five-year publication volume.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers--Elsevier,
Springer Nature, and Wiley--, which collectively accounted for 47\% of
hybrid journals and 62\% of articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358
artricles (33\% of the total) published across 1,951 hybrid
journals(16\% of the total). Springer Nature followed with 1,247,578
articles (17\%) in 2,311 hybrid journals (19\%), while Wiley recorded
858,939 articles (12\%) in1,382 hybrid journals (11\%). The remaining 54
publishers collectively accounted for 2,749,847 articles (38\%) in 6,476
hybrid journals (53\%).

@@

The three largest publishers--Elsevier, Springer Nature and Wiley--were
best represented in the exclusive intersection of all three data sources
(hoaddata, Scopus, and Web of Science). Together, they comprised 4,384
hybrid journals (55\% of the intersectional set) and dominated article
coverage with 4,174,315 (66\%), as determined through shared DOIs. When
examining publication volume per journal,
(\ref{fig-upset_coverage_results_publisher}C), Elsevier published, on
average, the largest journals, followed by Springer Nature and Wiley.

A comparative analysis of publisher portfolios across different indexing
sets demonstrates that publisher were not represented uniformly.
Springer Nature exhibited 519 hybrid journals exclusively indexed in
both hoaddata and Scopus. This set included journals from the Chinese
Academy of Science, German-language medical journals, and Eastern
European publications including the \emph{Journal of Mathematical
Sciences}, which also publishes English-language translations of
Russian-language works. Additionally, this subset included titles with a
more broader disciplianry focus such as \emph{SN Computer Science} and
newly launched hybrid journals like \emph{Nature Computational Science},
which started in 2021 and was indexed in Scopus but not yet in Web of
Science. The set also captured ceased journals, providing insights into
the dynamic nature of journal portfolios and business strategies.

Further examination of publisher portfolios not covered by hoaddata but
present in Scopus or Web of Science identified several publishers with
missing Creative Commons (CC) license information in Crossref. In
particular, Emerald represented 322 journals with 86,409 articles, AIP
Publishing accounted for 24 journals with 64,898 articles, and World
Scientific recorded 87 journals and 42,531 articles. In total, 9
publishers were not represented in Crossref.

An inspection of individual journals uncovered discrepancies in
Unpaywall's open access identification for certain publishers that
typically deposit CC licenses with Crossref. Notably, some
subscription-only journals contained one or two articles erroneously
tagged as hybrid open access by Unpaywall, which were subsequently
reflected in Scopus and Web of Science. Examples of such
misclassifications include Elsevier's \emph{Journal of Bioscience and
Bioengineering} and Springer Nature's \emph{Journal of Mechanical
Science and Technology}.

@@

This section examines the uptake of open access in hybrid journals,
focusing on the influence of transformative agreements across hoaddata,
Scopus, and Web of Science. The analysis assesses whether consistent
results can be derived from these data sources despite differences in
coverage and methodologies. Following Jahn (2025), indicators were
calculated for each data source and comprise the number and proportion
of open access articles, including those enabled by transformative
agreements, from 2019 to 2023. For Web of Science and Scopus, the impact
of transformative agreements was estimated using both first and
corresponding authorships, while hoaddata indicator calculation was
limited to first-author affiliations.

@@

Figure \ref{fig-uptake_overview}A shows a moderate growth of open access
in hybrid journals, which is consistent across hoaddata (black line),
Scopus (orange line), and Web of Science (blue line). According to
hoaddata, hybrid open access uptake increased from 6.4\% (n = 85,071) in
2019 to 19\% (n = 302,358) in 2023. Similarly, Scopus recorded an growth
from 6.5\% (n = 84,648) in 2019 to 19\% (n = 322,850) in 2023. However,
Web of Science recorded higher open access uptake in early years, before
converging in 2023, from 11\% (n = 137,202) in 2019 to 20\% (n =
255,481) in 2023. This may reflect delayed open access being labeled as
hybrid in Web of Science.

Open access by transformative agreements increased between 2019 and 2023
(Figures \ref{fig-uptake_overview}B and C). Trends were consistent for
first (dashed line) and corresponding author (dotted line) affiliations.
According to Scopus, 479,297 open access articles could be attributed to
transformative agreements based on first author metadata (increasing
from 15,341 to 206,084) and 489,262 using corresponding author metadata
(from 15,444 to 210,816). Web of Science recorded 493,028 open access
articles via transformative agreements using first author metadata
(increasing from 21,871 to 174,126) and 500,076 using corresponding
author metadata (from 22,092 to 177,030). hoaddata, lacking
corresponding author data, linked 501,649 articles to transformative
agreements but showed slower growth than Scopus and Web of Science, from
16,010 to 183,757.

From 2021 (hoadata, Scopus) resp. 2022 (Web of Science), transformative
agreements enabled the majority of hybrid open access. For first
authors, the share ranged between 68 \% (Web of Science), 64\% (Scopus),
and 61\% (hoaddata) in 2023. For corresponding authors, the shares were
slightly larger, with Web of Science recording 69 \% and Scopus 65\% in
2023. However, substantial hybrid open access was still facilitated
outside transformative agreements, likely through APCs payed from
discretionary research funds (ref).

@@

When considering open access trends by publisher (see
\ref{fig-uptake_publisher}), the observed differences in early uptake
rates between hoaddata and Scopus compared to Web of Science can be
largely attributed to articles published in Elsevier hybrid journals,
the largest publisher in our sample. Both hoadata and Scopus reported an
steady increase in open access uptake between 2019 and 2023 (hoaddata
from 4\% to 13\%; Scopus from 5\% to 15\%). In contrast, Elsevier's
share remained relatively constant, increasing only slightly from 14\%
to 15\% according to the Web of Science. Upon inspection, this
discrepancy primarily stemmed from articles provided under Elsevier's
user license, used for the publisher's open archive. These articles,
made freely available after an embargo period, were tagged as hybrid
open access in Web of Science, even though its documentation specified
that only articles under a CC license variant were considered. Previous
research (Haustein et al., 2024; Jahn et al., 2022) has shown that
Elsevier provided a substantial portion of its articles under this
license, explaining the relatively large and stable share of open access
over the years.

>>>>

This study investigates the suitability of open scholarly data sources
for assessing the impact of transformative agreements on hybrid open
access. As shown in Figure \ref{fig:workflow}, the methodology involved
comparing hoaddata, an openly available collection of open research
information on hybrid open access, with the bibliometric databases Web
of Science and Scopus. This section introduces the initial data sources,
followed by a presentation of the necessary data processing steps to
obtain eligible articles enabled by transformative agreements using
author roles (first and corresponding) and harmonised affiliation data.

@@

hoaddata, developed and maintained by the author to support open access
monitoring and research (Jahn, 2025), is an R data package collecting
information on hybrid open access uptake from multiple openly available
data sources (Jahn, 2024). It combines article-level metadata from
Crossref and OpenAlex with transformative agreement information from the
cOAlition S Journal Checker Tool (JCT), which links journal and
institutional data to agreements in the ESAC registry.

As discussed elsewhere (Jahn, 2025), hoaddata uses a unified dataset of
various safeguarded JCT snapshots. Journal data were enriched with ISSN
variants linked to an ISSN-L and institution data with associated
ROR-IDs representing university hospitals or institutes of large
research organisations, according to OpenAlex' institution entity. A
comprehensive exclusion of fully open access journals using multiple
journal lists (DOAJ, OpenAlex, Bielefeld) was performed to identify
hybrid journals. hoaddata relies on Crossref for obtaining journal
publication volume and open access status through Creative Commons
licence information relative to the published version (``version of
record''). Because of limited affiliation metadata in Crossref (Eck \&
Waltman, 2022), hoaddata sources affiliation metadata from OpenAlex.

hoaddata follows good practices for computational reproducibility using
R (Marwick et al., 2018). The package, which includes data, code, a test
suite and documentation, is openly available on GitHub. To ensure
computational reproducibility while aggregating the data, a GitHub
Actions continuous integration and delivery (CI/CD) workflow handles
data retrieval from the SUB Göttingen's open scholarly data warehouse
based on Google BigQuery, which provides high-performant programmatic
access to monthly snapshots of Crossref and OpenAlex.\footnote{\url{https://subugoe.github.io/scholcomm_analytics/data.html}}
The workflow has run regularly to fetch updates from these data sources
since 2022. The package version used in this study is 0.3, containing
data from the Crossref 2024-08 dump provided to Crossref Metadata Plus
subscribers and the OpenAlex 2024-08-29 monthly dump. It covers
agreements collected between July 2021 to July 2024 from the JCT. This
version including the computation log is available on GitHub
(https://github.com/subugoe/hoaddata/releases/tag/v.0.3).

@@

Clarivate Analytics' Web of Science (WoS) is a well-established
proprietary bibliometric database consisting of several collections
(Birkle et al., 2020). The collections considered in this study were the
Science Citation Index Expanded (SCIE), the Social Sciences Citation
Index (SSCI) and the Arts \& Humanities Citation Index (AHCI).

These collections provides important data points for analysing open
access: author affiliations and roles, differentiation of journal
articles into document types representing different types of journal
contributions, such as original articles or reviews, and open access
status information derived from OurResearch's Unpaywall, the same
provider as OpenAlex. However, the Web of Science lacks information
about journals and articles under transformative agreements.

For programmatic access to article-level data, this study used the
database of the Kompetenznetzwerk Bibliometrie (KB) in Germany. The KB
processes raw XML data provided by Clarivate Analytics, which is
ingested into an in-house PostgreSQL database under a uniform schema. To
support reproducibility, KB maintains annual snapshots of the database.
Accordingly, this study used the annual snapshot from April 2024
(wos\_b\_202404), which is considered to cover almost the entire
previous publication year (Schmidt et al., 2024).

@@

Using an enriched and unified list of hybrid journals in transformative
agreements according to the JCT, article metadata were retrieved from
each database using all ISSN variants linked to an ISSN-L.

The article metadata included DOIs, publication dates, open access
information as well as author roles and institutional affiliations.
Publication years were determined using the earliest known date of
publication in a journal. In hoaddata, this corresponded to Crossref's
issued date. For Web of Science and Scopus, the earliest publication
date was used where available, with Scopus dates specifically determined
by the KB through version tracking of the raw data.

Many transformative agreements typically cover only certain types of
journal articles, in particular original research articles including
reviews. Because of limited information on these document types in open
scholarly data (Haupka et al., 2024), hoaddata used an extended version
of Unpaywall's paratext recognition approach to exclude non-scholarly
content (Piwowar et al., 2018). To exclude conference supplements, which
are also often not covered by transformative agreements, only articles
published in regular issues, indicated by numerical pagination, were
considered. For Web of Science and Scopus, their established document
type classifications were used to identify original research articles
and reviews, referred to as original articles throughout this study.

@@

Articles in hybrid journals were considered open access when they were
made freely available under a Creative Commons license on publishers'
platforms. While hoaddata sourced this information from Crossref license
metadata, Web of Science and Scopus relied on Unpaywall as evidence
source. Unpaywall supplements Crossref metadata by parsing publisher
websites directly, addressing cases where publishers do not provide
machine-readable Creative Commons license information (Piwowar et al.,
2018). This additional parsing remains necessary despite transformative
agreement workflows recommend the depositing license information during
DOI registration (Geschuhn \& Stone, 2017). Both Web of Science and
Scopus defined hybrid open access consistently as content available
under Creative Commons licenses on publisher platforms, distinguishing
it from bronze open access that lack such explicit license information,
or use publisher-specific licenses.

@@

Author affiliations were retrieved for both first and, if available,
corresponding authors to prepare the linking between articles and
institutions covered by transformative agreements. To handle different
address variants, database-specific affiliation identifiers were used:
ROR-IDs from OpenAlex for hoaddata, affiliation enhanced names for Web
of Science, and Scopus Affiliation Identifiers for Scopus. Additionally,
ISO country codes were retrieved for each author's address to compile
country-level statistics. These country codes for Web of Science and
Scopus were provided by the KB.

Because neither Web of Science nor Scopus support ROR IDs, the
institution identifier used by the JCT, a two-step matching process was
implemented to harmonise affiliation data. First, 2,782,540 articles
from 6,457 institutions with ROR-IDs in the JCT data since 2017
(according to hoaddata) were processed to map first authors' ROR-IDs to
corresponding proprietary affiliation identifier in Web of Science and
Scopus using DOI matching. Then, an algorithm selected the most frequent
ROR ID and proprietary identifier pairs to handle multiple affiliations
and organisational hierarchy differences.

@@

As a result of the comprehensive data processing described above,
datasets on open access in hybrid journals included in transformative
agreements were aggregated for each database at country and journal
level by year. Table~\ref{tbl-methods_overview_tab} provides a general
overview of the coverage between 2019 and 2023. It shows that the
majority of hybrid journals published at least one original research
article or review marked as original during the five-year period. These
journals formed the basis for the subsequent calculation of
article-level indicators.

@@

\centering
\begin{tabular}[t]{lrrr}
\toprule
\textbf{} & \textbf{HOAD} & \textbf{Web of Science} & \textbf{Scopus}\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Hybrid journal coverage}}\\
\hspace{1em}Active journals & 12,890 & 8,655 & 11,888\\
\hspace{1em}Active journals (core) & 12,888 & 8,655 & 11,878\\
\hspace{1em}Active journals (core) with OA & 11,348 & 8,392 & 11,313\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Publication volume}}\\
\hspace{1em}Total published articles & 9,740,015 & 8,616,053 & 8,117,644\\
\hspace{1em}Original articles & 8,158,425 & 6,708,083 & 7,317,703\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Digital Object Identifier (DOI) coverage}}\\
\hspace{1em}Articles with DOI & 9,740,015 & 7,713,796 & 8,105,112\\
\hspace{1em}Original articles with DOI & 8,158,425 & 6,695,661 & 7,314,327\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Open Access (OA) metrics}}\\
\hspace{1em}OA articles & 998,699 & 1,112,758 & 974,099\\
\hspace{1em}Original OA articles & 969,817 & 1,019,784 & 922,578\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Original articles with affiliation data}}\\
\hspace{1em}First author articles & 7,242,542 & 6,294,855 & 7,232,017\\
\hspace{1em}Corresponding author articles & 5,534,207 & 6,291,441 & 6,898,487\\
\bottomrule
\end{tabular}

@@

While hoaddata only covers articles with a DOI, Scopus and Web of
Science publication indicators were calculated using their database
identifier. A subsequent comparison of DOI coverage shows that
non-original articles in Web of Science often lacked a DOI. This was
particularly the case for meeting abstracts, which are notably prevalent
in Health Sciences journals (Melero-Fuentes et al., 2025) and are not
indexed by Scopus (Donner, 2017). Open access indicators were aggregated
by DOI, as Unpaywall only collects information on open access status for
articles with a DOI. A closer look at original articles with affiliation
data reveals a lack of data on corresponding authors in the case of
OpenAlex, the affiliation data source used by hoaddata, compared to Web
of Science and Scopus. Only about two-thirds of the articles examined
provided corresponding author affiliations. For first authors, the
proportion was \%. Therefore, only first author data for hoaddata were
considered in the following analysis.

@@

Journal coverage analysis revealed that 66\% (n = 7,970) of hybrid
journals included in transformative agreements were indexed in all three
databases (Figure \ref{fig-upset_coverage_results}A). The second-largest
set consisted of journals indexed in both hoaddata and Scopus,
comprising 21\% (n = 2,595) of hybrid journals. Notably, 6\% (n = 739)
of journals were exclusively contained in hoaadata, while another 6\% (n
= 748) were only found in the proprietory database Scopus. Of these, 354
were also available in the Web of Science. Upon inspection, this group
of hybrid journals exclusively covered in the proprietary data sources
mainly represented hybrid journals for which no open access evidence
could be retrieved from Crossref, the open access evidence source for
hoaddata.

In terms of article coverage, Figure \ref{fig-upset_coverage_results}B
shows the total publication volume per combination in terms of DOI
availability. The largest set of hybrid journals, which includes all
three data sources, also contains the largest number of articles. In
total, these journals recorded 6,289,687 overlapping articles,
represented by the blue bar. They represented \% of original articles
with DOI indexed in the Web of Science sample, and \% in Scopus. Another
657,697 articles were exclusive to both Scopus and hoaddata. Exclusively
in hoaddata were 177,110 articles, and exclusively in the proprietary
databases were 325,194 articles.

Figure \ref{fig-upset_coverage_results}B also shows the surplus of
articles with DOI that were only available via hoaddata (grey area). In
case of hybrid journals covered by all three data sources, 1,023,882
DOIs were only present in hoaddata. After validation at the DOI level
using the KB databases and manual inspection, the main reasons for
missing DOI coverage in the proprietary database were insufficient
classification of journal content as original research articles and
reviews during the compilation of hoaddata. Particularly, letters and
editorials could not be fully detected. Paratext recognition failed in
37\% of DOIs to identify non-scholarly content such as front matters or
reviewer lists, which are generally not indexed by Scopus and Web of
Science. To a lesser extent, differences in publication and indexing
dates were a reason for non-overlapping DOIs.

Using overlapping DOIs, the publication volume between 2019 and 2023 was
also calculated for each journal. Figure
\ref{fig-upset_coverage_results}C illustrates the distribution for each
combination. It shows that there is a large spread across the journals
covered by all three data sources. Furthermore, journals in this set
published more on average than journals covered by only one or two data
sources. In particular, the journals covered exclusively by hoaddata
were substantially smaller than those covered by all three sources. Upon
inspection, these were often newly launched journals, which explains the
relatively low five-year publication volume. An example is \emph{Digital
society} that published 86 articles. This hybrid journal was launched in
2022, being covered by various Springer Nature transformative agreements
since then.

@@

Figure \ref{fig-upset_coverage_results_publisher} presents the coverage
of hybrid journals in transformative agreements across data sources from
2019 to 2023 with a focus on publisher portfolios. The analysis
highlights the dominance of the three largest publishers--Elsevier,
Springer Nature, and Wiley--, which collectively accounted for 47\% of
hybrid journals and 62\% of articles published during this five-year
period. In terms of article volume, Elsevier led with 2,441,358 articles
(33\% of the total) published across 1,951 hybrid journals (16\% of the
total). Springer Nature followed with 1,247,578 articles (17\%) in
2,311, although recording the largest number of hybrid journals (19\%).
Wiley accounted for 858,939 articles (12\%) in1,382 hybrid journals
(11\%). The remaining 54 publishers collectively accounted for 2,749,847
articles (38\%) in 6,476 hybrid journals (53\%).

@@

The three largest publishers--Elsevier, Springer Nature and Wiley--were
best represented in the exclusive intersection of all three data sources
(hoaddata, Scopus, and Web of Science). Together, they comprised 4,384
hybrid journals (55\% of the intersectional set) and dominated article
coverage (n = 4,174,315; 66\%), as determined through shared DOIs. When
examining publication volume per journal, (Figure
\ref{fig-upset_coverage_results_publisher}C), Elsevier published, on
average, the largest journals, followed by Springer Nature and Wiley.

A comparative analysis of publisher portfolios across different indexing
sets demonstrates that publisher were not represented uniformly.
Notably, Springer Nature exhibited 519 hybrid journals exclusively
indexed in both hoaddata and Scopus. This set included journals from the
Chinese Academy of Science, German-language medical journals, and
Eastern European publications including the \emph{Journal of
Mathematical Sciences}, which also publishes English-language
translations of Russian-language works. Additionally, this subset
included titles with a more broader disciplinary focus such as \emph{SN
Computer Science} and newly launched hybrid journals like \emph{Nature
Computational Science}, which started in 2021 and was indexed in Scopus
but not yet in Web of Science. The set also captured ceased journals,
providing further insights into the dynamic nature of journal portfolios
and business strategies.

Examining publisher portfolios not covered by hoaddata but present in
Scopus or Web of Science identified several publishers with missing CC
license information in Crossref. In particular, Emerald represented 322
journals with 86,409 articles, AIP Publishing accounted for 24 journals
with 64,898 articles, and World Scientific recorded 87 journals and
42,531 articles. In total, 9 publishers were not represented in
Crossref.

An inspection of individual journals also uncovered discrepancies in
Unpaywall's open access identification for certain publishers that
typically deposit CC licenses with Crossref. Notably, some
subscription-only journals contained one or two articles erroneously
tagged as hybrid open access by Unpaywall, which were subsequently
reflected in Scopus and Web of Science. Examples of such
misclassifications include Elsevier's \emph{Journal of Bioscience and
Bioengineering} and Springer Nature's \emph{Journal of Mechanical
Science and Technology}.

@@

This section examines the uptake of open access in hybrid journals,
focusing on the influence of transformative agreements across hoaddata,
Scopus, and Web of Science. The aim was to assess whether consistent
results can be derived from these data sources despite differences in
coverage and methodologies. Following Jahn (2025), indicators were
calculated for each data source and comprise the number and proportion
of open access articles, including those enabled by transformative
agreements, from 2019 to 2023. For Web of Science and Scopus, the impact
of transformative agreements was estimated using both first and
corresponding authorships, while hoaddata indicator calculation was
limited to first-author affiliations.

@@

Figure \ref{fig-uptake_overview}A shows a moderate growth of open access
in hybrid journals, which is consistent across hoaddata (black line),
Scopus (orange line), and Web of Science (blue line). According to
hoaddata, hybrid open access uptake increased from 6.4\% (n = 85,071) in
2019 to 19\% (n = 302,358) in 2023. Similarly, Scopus recorded an growth
from 6.5\% (n = 84,648) in 2019 to 19\% (n = 322,850) in 2023. However,
Web of Science recorded higher open access uptake in early years, before
converging in 2023, from 11\% (n = 137,202) in 2019 to 20\% (n =
255,481) in 2023,suggesting an different approach towards labeling
hybrid open access by the Web of Science.

Simialry, hybrid open access by transformative agreements increased
between 2019 and 2023 (Figures \ref{fig-uptake_overview}B and C). Trends
were consistent for first (dashed line) and corresponding author (dotted
line) affiliations. According to Scopus, 479,297 open access articles
could be attributed to transformative agreements based on first author
metadata (increasing from 15,341 to 206,084) and 489,262 using
corresponding author metadata (from 15,444 to 210,816). Web of Science
recorded 493,028 open access articles via transformative agreements
using first author metadata (increasing from 21,871 to 174,126) and
500,076 using corresponding author metadata (from 22,092 to 177,030).
hoaddata, lacking corresponding author data, linked 501,649 articles to
transformative agreements but showed slower growth than Scopus and Web
of Science, from 16,010 to 183,757.

From 2021 (hoadata, Scopus) resp. 2022 (Web of Science), transformative
agreements enabled the majority of hybrid open access. For first
authors, the share ranged between 61\% (hoaddata), 64\% (Scopus), and 68
\% (Web of Science) in 2023. For corresponding authors, the shares were
slightly larger, with Scopus recording 65\% and Web of Science 69\% in
2023. However, substantial hybrid open access was still facilitated
outside transformative agreements, likely through APCs payed from
discretionary research funds (ref).

@@

When considering open access trends by publisher (see Figure
\ref{fig-uptake_publisher}), the observed differences in early uptake
rates between hoaddata and Scopus compared to Web of Science can be
largely attributed to articles published in Elsevier hybrid journals,
the largest publisher in our sample. Both hoadata and Scopus reported an
steady increase in open access uptake between 2019 and 2023 (hoaddata
from 4\% to 13\%; Scopus from 5\% to 15\%). In contrast, Elsevier's
share remained relatively constant, increasing only slightly from 14\%
to 15\% according to the Web of Science. Upon inspection, this
discrepancy primarily stemmed from articles provided under Elsevier's
user license, used for the publisher's open archive. These articles,
made freely available after an embargo period, were tagged as hybrid
open access in Web of Science, even though its documentation specified
that only articles under a CC license variant were considered. Previous
research (Haustein et al., 2024; Jahn et al., 2022) has shown that
Elsevier provided a substantial portion of its articles under this
license, explaining the relatively large and stable share of open access
over the years.
